Formalized Linear Programming

### 2019-12-31 TeX/paper.tex
Correct variable names. There were two different things called c

\section{Linear Programming}
\chris{The standard simplex algorithm. No reference to Lean code}
The simplex algorithm aims to optimize the value of an objective function within the constraints of a set of affine inequalities. It can be used to solve problems of the following form.
% \color{red} This is a slightly nonstandard way of writing a linear program. The usual way would be $Ax \le b$. However the matrix that is stored during a simplex is $-A$ according to that representation, but $A$ according to the representation below.\color{black}
\begin{equation}
  \begin{aligned}
    \text{Find } x \in \mathbb{Q}^n \text{ that maximizes } && c^Tx \\
    \text{subject to } && A x + b \ge 0\\
    \text{and } && x \ge 0
  \end{aligned}
\end{equation}
Here, $x \in \mathbb{Q}^n$, $c \in \mathbb{Q}^n$, $b \in \mathbb{Q}^m$, and $A \in \mathbb{Q}^{m \times n}$.
This can be written in an equivalent form \chris{The inclusion of $z$ in the objective function is slightly unusual because obviously maximizing $c^Tx$ is the same as maximizing $c^Tx + z$. However $z$ is a cell in the tableau, it is the objective function value of the sample solution, and the explanation of the tableau is slightly easier with it}
\begin{equation}
  \begin{aligned}
    \text{Find } x_r \text{ and } x_c \text{ that maximize } && c^Tx_c + z \\
    \text{subject to } && x_r = A x_c + b \\
    \text{and } && x_r \ge 0, x_c \ge 0
  \end{aligned}
\end{equation}
with $x_r \in \mathbb{Q}^m$, $x_c \in \mathbb{Q}^n$ and $z \in \mathbb{Q}$

@@

In the tableau representation of this problem, the vectors $x_r$ and $x_c$ are viewed as minors of a single variable $x$. The subset of $\mathbb{Q}^{m+n}$, $\{x : x_c = Ax_r + b \wedge x_c \ge 0 \wedge x_r \ge 0\}$ is stored as part of the simplex algorithm. The tableau consists of a matrix $A$ of coefficients, a column vector $b$, a reduced cost vector $c$, and the constant part of the objective function $z$. Additionally two vectors of indices $r$ and $c$ defining the minors $x_r$ and $x_c$ are also stored.
\begin{equation}
  \begin{aligned}
    x_{r_1} && = && A_{11}x_{c_1} && + && A_{12}x_{c_2} && + && \dots && + && A_{1n}x_{c_n} && + && b_1 \\
    x_{r_2} && = && A_{21}x_{c_1} && + && A_{22}x_{c_2} && + && \dots && + && A_{2n}x_{c_n} && + && b_2 \\
    \vdots && && \vdots && && \vdots && && && && \vdots && \vdots \\
    x_{r_m} && = && A_{m1}x_{c_1} && + && A_{m2}x_{c_2} && + && \dots && + && A_{mn}x_{c_n} && + && b_n \\
    c^Tx_c + z && = && c_1x_{c_1} && + && c_2x_{c_2} && + && \dots && + && c_nx_{c_n} && + && z
  \end{aligned}
\end{equation}
In this tableau representation the $x_r$ are known as the \textit{row} variables, and the $x_c$ are the \textit{column} variables. In a tableau the row variables are expressed as affine functions in terms of the column variables.

The simplex algorithm iteratively performs the \textit{pivot} operation on a tableau. The pivot operation takes a row and a column index as an argument, and swaps their positions in the tableau. If $x_{r_i}$ and $x_{c_j}$ are pivoted, then in the new tableau $x_{r_i}$ becomes a row variable and $x_{c_j}$ becomes a column variable. The variables $(x_{r_1}, x_{r_2}, \dots, x_{c_j}, \dots x_{r_m})$ are expressed as affine functions of $(x_{c_1}, x_{c_2}, \dots, x_{r_i}, \dots, x_{c_n})$. The matrix $A$, and the vectors $c$ and $b$ are updated to $A'$, $c'$ and $b'$, such that the tableau represents the same subset of $\mathbb{Q}^{m + n}$, and the same optimization problem. Additionally the vectors $r$ and $c$ are updated to vectors $r' := (r_1, r_2, \dots c_j, \dots, r_m)$ and $c' := (c_1, c_2 \dots r_i, \dots, c_n)$.
\chris{Oops, two things are called $c$} Preserving the same polyhedron is possible if $A_{ij} \ne 0$.
\begin{equation}
  \begin{aligned}
    x_{r_1} && = && A'_{11}x_{c_1} && + && A'_{12}x_{c_2} && + && \dots && + && A'_{1j}x_{r_i} && + && \dots && + && A'_{1n}x_{c_n} && + && b'_1\\
    x_{r_2} && = && A'_{21}x_{c_1} && + && A'_{22}x_{c_2} && + && \dots && + && A'_{2j}x_{r_i} && + && \dots && + && A'_{2n}x_{c_n} && + && b'_2\\
    \vdots && && \vdots && && \vdots && && && && \vdots && && && && \vdots && && \vdots \\
    x_{c_j} && = && A'_{i1}x_{c_1} && + && A'_{i2}x_{c_2} && + && \dots && + && A'_{ij}x_{r_i}&& + && \dots && + && A'_{in}x_{c_n} && + && b'_i\\
    \vdots && && \vdots && && \vdots && && && && \vdots && && && && \vdots && && \vdots \\
    x_{r_m} && = && A'_{m1}x_{c_1} && + && A'_{m2}x_{c_2} && + && \dots && + && A'_{mj}x_{r_i}&& + && \dots && + && A'_{mn}x_{c_n} && + && b'_n\\
    c^Tx + z && = && c'_1x_{c_1} && + && c'_2x_{c_2} && + && \dots && + && c'_jx_{r_i} && + && \dots && + && c'_nx_{c_n} && + && z'
  \end{aligned}
\end{equation}
The expressions for the updated matrix $A'$, and the updated vectors $c'$, $b'$ are given by the following formulas \cite{Nelson:1980:TPV:909447}.
\begin{equation}\label{eqn:pivot_tableau}
A'_{i'j'} = \begin{cases}
  A_{ij}^{-1} & \text{if } i' = i \text{ and } j' = j \\
  -A_{ij'}/A_{ij} & \text{if } i' = i \text{ and } j' \ne j \\
  A_{i'j}/A_{ij} & \text{if } i' \ne i \text{ and } j' = j \\
  A_{i'j'} - A_{i'j}A_{ij'}/A_{ij}  & \text{if } i' \ne i \text{ and } j' \ne j
\end{cases}
\end{equation}
\begin{equation}\label{eqn:pivot_const}
  b'_{i'} = \begin{cases}
    -b_i/A_{ij} & \text{if } i' = i \\
    b_{i'} - A_{i'j}b_i/A_{ij} & \text{if } i' \ne i
  \end{cases}
\end{equation}
\begin{equation}
  c'_{j'} = \begin{cases}
    c_j/A_{ij} & \text{if } j' = j \\
    c_{j'} - c_jA_{ij'}/A_{ij} & \text{if } j' \ne j
  \end{cases}
\end{equation}
\begin{equation}\label{eqn:pivot_obj}
  z' = z - c_jb_i/A_{ij}
\end{equation}

@@

Given any assignment of values to the column variables of a tableau, there is a unique $x$ satisfying $x_r=Ax_c+b$ The \textit{sample solution} of a tableau is the solution of $x_r = Ax_c+b$ found by setting $x_c$ to zero. For this solution $x_r=b$. Similarly, the objective function must have value $z$ in the \textit{sample solution}. A tableau is \textit{feasible} if the sample solution also satisfies $x \ge 0$. This is equivalent to saying $b \ge 0$.

@@

If there is no column $j$ satisfying $c_j > 0$, then the sample solution must maximize $c^Tx_c$. This is because $c^T \le 0$ and $x_c \ge 0$, so $c^Tx_c \le 0$, and the maximum value of this is given by setting $x_c$ to zero.

@@

\chris{Our version, dead columns, unrestricted variables, stopping early. Also very little reference to Lean code.}
The implementation in Lean follows the implemenentation in the Simplify prover \cite{simplify}. This implementation represents a polyhedron in a slightly different way. There are two additional sets stored, a set of \textit{restricted} variables and a set of \textit{dead} columns.
The optimization problem for this implementation is given by
\begin{equation}
  \begin{aligned}
  \text{Find } x \text{ that maximizes } && x_{r_k} \\
  \text{such that } && x_r = A x_c + b \\
  \text{and } && \forall i \in restricted, x_i \ge 0 \\
  \text{and } && \forall j \in dead, x_{c_j} = 0
  \end{aligned}
\end{equation}

@@

\begin{equation}
  \begin{aligned}
    A_{kj} \ne 0 \\
    c_j \in restricted \implies A_{kj} > 0 \\
    j \notin dead \\
    i \ne k \\
    r_i \in restricted \\
    A_{ij} / A_{kj} < 0 \\
    \forall i', r_{i'} \in restricted \wedge A_{i'j} / A_{kj} < 0 \implies \left|b_i/A_{ij}\right| \le \left|b_{i'} / A_{i'j}\right|
  \end{aligned}
\end{equation}

@@

Suppose $R^T$ is the matrix mapping $x$ to $x_r$, and $C^T$ is the matrix mapping $x$ to $x_c$. Then $R$ and $C$ have the following properties.
\begin{equation}\label{eqn:pequiv_props}
  \begin{aligned}
    R^TR = 1 \\
    C^TC = 1 \\
    R^TC = 0 \\
    C^TR = 0 \\
    RR^T + CC^T = 1
  \end{aligned}
\end{equation}
These matrices are used in the definitions of the polyhedron corresponding to a tableau, and also the definition of the function \lstinline|of_col| can be made short by always expressing minors using matrix multiplication. Given $x_c$, and a tableau, the function $ofcol$ computes $x$, satisfying $x_r = Ax_c + b$. The definition is given by
\begin{equation}
  ofcol(x_c) := Cx_c + R(Ax_c+b)
\end{equation}
In this case we are not using $C$ and $R$ to take minors, but rather using $C$ to assign values to the column variables, and leave zeros elsewhere, and $R$ to assign values to the row variables and zeros elsewhere. The basic properties of this function can be easily proved using the identities in (\ref{eqn:pequiv_props}).
\begin{equation}
  \begin{aligned}
    C^Tofcol(x_c) = x_c \\
    R^Tofcol(x_c) = Ax_c + b \\
    R^Tofcol(x_c) = AC^Tofcol(x_c) + b
  \end{aligned}
\end{equation}

@@

% The type \verb|fin (m + n)| is the type of nonnegative integers less than \verb|m + n|. The \verb|row_indices| and \verb|column_indices| fields of the structure are vectors of length \verb|m| and \verb|n| respectively, containing elements of \verb|fin (m + n)|. The final field of the structure stipulates that all element of \verb|fin (m + n)| are an element of either \verb|row_indices| or \verb|col_indices|. This also implies, via a pigeonhole argument, that the two vectors are disjoint and have no duplicates. Given vectors $r$ and $c$ of row and column indices, it is possible to define the following minors of any vector $x \in \mathbb{Q}^{m + n}$.
% \begin{equation}\label{eqn:minor}
% \begin{aligned}
% x_r := (x_{r_0}, ... , x_{r_{m-1}}) \\
% x_c := (x_{c_0}, ... , x_{c_{n-1}})
% \end{aligned}
% \end{equation}

@@

% These ``row'' variables are then expressed as affine combinations of the ``column'' variables by the tableau. Given an $m \times n$ matrix $A$, and a constant vector $k \in \mathbb{Q}^m$, we can express the following relation between the current row variables $x_r$ and column variables $x_c$.
% \begin{equation}\label{eqn:axk}
%   x_r = Ax_c + k
% \end{equation}

@@

% \verb|T.to_partition.rowp.to_matrix| is an $m \times (m + n)$ matrix, which corresponds to the linear map taking the minor $x_r$ of $x$ defined in (\ref{eqn:minor}). \verb|T.to_partition.colp.to_matrix| does the same for the column variables. This equation is the same as the equation in (\ref{eqn:axk}). The notation \lstinline|⬝| is used for matrix multiplication.

>>>>

\section{Linear Programming}
\chris{The standard simplex algorithm. No reference to Lean code}
The simplex algorithm aims to optimize the value of an objective function within the constraints of a set of affine inequalities. It can be used to solve problems of the following form.
% \color{red} This is a slightly nonstandard way of writing a linear program. The usual way would be $Ax \le b$. However the matrix that is stored during a simplex is $-A$ according to that representation, but $A$ according to the representation below.\color{black}
\begin{equation}
  \begin{aligned}
    \text{Find } x \in \mathbb{Q}^n \text{ that maximizes } && c^Tx \\
    \text{subject to } && A x + b \ge 0\\
    \text{and } && x \ge 0
  \end{aligned}
\end{equation}
Here, $x \in \mathbb{Q}^n$, $c \in \mathbb{Q}^n$, $b \in \mathbb{Q}^m$, and $A \in \mathbb{Q}^{m \times n}$.
This can be written in an equivalent form \chris{The inclusion of $z$ in the objective function is slightly unusual because obviously maximizing $c^Tx$ is the same as maximizing $c^Tx + z$. However $z$ is a cell in the tableau, it is the objective function value of the sample solution, and the explanation of the tableau is slightly easier with it}
\begin{equation}
  \begin{aligned}
    \text{Find } x_{row} \text{ and } x_{col} \text{ that maximize } && c^Tx_{col} + z \\
    \text{subject to } && x_{row} = A x_{col} + b \\
    \text{and } && x_{row} \ge 0, x_{col} \ge 0
  \end{aligned}
\end{equation}
with $x_{row} \in \mathbb{Q}^m$, $x_{col} \in \mathbb{Q}^n$ and $z \in \mathbb{Q}$

@@

In the tableau representation of this problem, the vectors $x_{row}$ and $x_{col}$ are viewed as minors of a single variable $x$. The subset of $\mathbb{Q}^{m+n}$, $\{x : x_{col} = Ax_{row} + b \wedge x_{col} \ge 0 \wedge x_{row} \ge 0\}$ is stored as part of the simplex algorithm. The tableau consists of a matrix $A$ of coefficients, a column vector $b$, a reduced cost vector $c$, and the constant part of the objective function $z$. Additionally two vectors of indices $row$ and $col$ defining the minors $x_{row}$ and $x_{col}$ are also stored.
The polyhedron defined by a tableau is given by the following set of equations.
\begin{equation}
  \begin{aligned}
    x_{row_1} && = && A_{11}x_{col_1} && + && A_{12}x_{col_2} && + && \dots && + && A_{1n}x_{col_n} && + && b_1 \\
    x_{row_2} && = && A_{21}x_{col_1} && + && A_{22}x_{col_2} && + && \dots && + && A_{2n}x_{col_n} && + && b_2 \\
    \vdots && && \vdots && && \vdots && && && && \vdots && \vdots \\
    x_{row_m} && = && A_{m1}x_{col_1} && + && A_{m2}x_{col_2} && + && \dots && + && A_{mn}x_{col_n} && + && b_n \\
    c^Tx_{col} + z && = && c_1x_{col_1} && + && c_2x_{col_2} && + && \dots && + && c_nx_{col_n} && + && z
  \end{aligned}
\end{equation}
In this tableau representation the $x_{row}$ are known as the \textit{row} variables, and the $x_{col}$ are the \textit{column} variables. In a tableau the row variables are expressed as affine functions in terms of the column variables.

The simplex algorithm iteratively performs the \textit{pivot} operation on a tableau. The pivot operation takes a row and a column index as an argument, and swaps their positions in the tableau. If $x_{row_i}$ and $x_{col_j}$ are pivoted, then in the new tableau $x_{row_i}$ becomes a row variable and $x_{col_j}$ becomes a column variable. The variables $(x_{row_1}, x_{row_2}, \dots, x_{col_j}, \dots x_{row_m})$ are expressed as affine functions of $(x_{col_1}, x_{col_2}, \dots, x_{row_i}, \dots, x_{col_n})$. The matrix $A$, and the vectors $c$ and $b$ are updated to $A'$, $c'$ and $b'$, such that the tableau represents the same subset of $\mathbb{Q}^{m + n}$, and the same optimization problem.
Additionally, the vectors $row$ and $col$ are updated to vectors $row' := (row_1, row_2, \dots col_j, \dots, row_m)$ and $col' := (col_1, col_2 \dots row_i, \dots, col_n)$.
\chris{Oops, two things are called $c$} Preserving the same polyhedron is possible if $A_{ij} \ne 0$.
\begin{equation}
  \begin{aligned}
    x_{row_1} && = && A'_{11}x_{col_1} && + && A'_{12}x_{col_2} && + && \dots && + && A'_{1j}x_{row_i} && + && \dots && + && A'_{1n}x_{col_n} && + && b'_1\\
    x_{row_2} && = && A'_{21}x_{col_1} && + && A'_{22}x_{col_2} && + && \dots && + && A'_{2j}x_{row_i} && + && \dots && + && A'_{2n}x_{col_n} && + && b'_2\\
    \vdots && && \vdots && && \vdots && && && && \vdots && && && && \vdots && && \vdots \\
    x_{col_j} && = && A'_{i1}x_{col_1} && + && A'_{i2}x_{col_2} && + && \dots && + && A'_{ij}x_{row_i}&& + && \dots && + && A'_{in}x_{col_n} && + && b'_i\\
    \vdots && && \vdots && && \vdots && && && && \vdots && && && && \vdots && && \vdots \\
    x_{row_m} && = && A'_{m1}x_{col_1} && + && A'_{m2}x_{col_2} && + && \dots && + && A'_{mj}x_{row_i}&& + && \dots && + && A'_{mn}x_{col_n} && + && b'_n\\
    c^Tx + z && = && c'_1x_{col_1} && + && c'_2x_{col_2} && + && \dots && + && c'_jx_{row_i} && + && \dots && + && c'_nx_{col_n} && + && z'
  \end{aligned}
\end{equation}
The expressions for the updated matrix $A'$, and the updated vectors $c'$, $b'$ are given by the following formulas \cite{Nelson:1980:TPV:909447}.
\begin{equation}\label{eqn:pivot_tableau}
A'_{i'j'} = \begin{cases}
  A_{ij}^{-1} & \text{if } i' = i \text{ and } j' = j \\
  -A_{ij'}/A_{ij} & \text{if } i' = i \text{ and } j' \ne j \\
  A_{i'j}/A_{ij} & \text{if } i' \ne i \text{ and } j' = j \\
  A_{i'j'} - A_{i'j}A_{ij'}/A_{ij}  & \text{if } i' \ne i \text{ and } j' \ne j
\end{cases}
\end{equation}
\begin{equation}\label{eqn:pivot_const}
  b'_{i'} = \begin{cases}
    -b_i/A_{ij} & \text{if } i' = i \\
    b_{i'} - A_{i'j}b_i/A_{ij} & \text{if } i' \ne i
  \end{cases}
\end{equation}
\begin{equation}
  c'_{j'} = \begin{cases}
    c_j/A_{ij} & \text{if } j' = j \\
    c_{j'} - c_jA_{ij'}/A_{ij} & \text{if } j' \ne j
  \end{cases}
\end{equation}
\begin{equation}\label{eqn:pivot_obj}
  z' = z - c_jb_i/A_{ij}
\end{equation}

@@

Given any assignment of values to the column variables of a tableau, there is a unique $x$ satisfying $x_{row}=Ax_{col}+b$ The \textit{sample solution} of a tableau is the solution of $x_{row} = Ax_{col}+b$ found by setting $x_{col}$ to zero. For this solution $x_{row}=b$. Similarly, the objective function must have value $z$ in the \textit{sample solution}. A tableau is \textit{feasible} if the sample solution also satisfies $x \ge 0$. This is equivalent to saying $b \ge 0$.

@@

If there is no column $j$ satisfying $c_j > 0$, then the sample solution must maximize $c^Tx_{col}$. This is because $c^T \le 0$ and $x_{col} \ge 0$, so $c^Tx_{col} \le 0$, and the maximum value of this is given by setting $x_{col}$ to zero.

@@

\chris{Our version, dead columns, unrestricted variables, stopping early. Also very little reference to Lean code.}
The implementation in Lean follows the implemenentation in the Simplify prover \cite{simplify}. This implementation represents a polyhedron in a slightly different way. There are two additional sets stored, a set of \textit{restricted} variables and a set of \textit{dead} columns.
The optimization problem for this implementation is given by
\begin{equation}
  \begin{aligned}
  \text{Find } x \text{ that maximizes } && x_{row_k} \\
  \text{such that } && x_{row} = A x_{col} + b \\
  \text{and } && \forall i \in restricted, x_i \ge 0 \\
  \text{and } && \forall j \in dead, x_{col_j} = 0
  \end{aligned}
\end{equation}

@@

\begin{equation}
  \begin{aligned}
    A_{kj} \ne 0 \\
    col_j \in restricted \implies A_{kj} > 0 \\
    j \notin dead \\
    i \ne k \\
    row_i \in restricted \\
    A_{ij} / A_{kj} < 0 \\
    \forall i', row_{i'} \in restricted \wedge A_{i'j} / A_{kj} < 0 \implies \left|b_i/A_{ij}\right| \le \left|b_{i'} / A_{i'j}\right|
  \end{aligned}
\end{equation}

@@

Suppose $R^T$ is the matrix mapping $x$ to $x_{row}$, and $C^T$ is the matrix mapping $x$ to $x_{col}$. Then $R$ and $C$ have the following properties.
\begin{equation}\label{eqn:pequiv_props}
  \begin{aligned}
    R^TR = 1 \\
    C^TC = 1 \\
    R^TC = 0 \\
    C^TR = 0 \\
    RR^T + CC^T = 1
  \end{aligned}
\end{equation}
These matrices are used in the definitions of the polyhedron corresponding to a tableau, and also the definition of the function \lstinline|of_col| can be made short by always expressing minors using matrix multiplication. Given $x_{col}$, and a tableau, the function $ofcol$ computes $x$, satisfying $x_{row} = Ax_{col} + b$. The definition is given by
\begin{equation}
  ofcol(x_{col}) := Cx_{col} + R(Ax_{col}+b)
\end{equation}
In this case we are not using $C$ and $R$ to take minors, but rather using $C$ to assign values to the column variables, and leave zeros elsewhere, and $R$ to assign values to the row variables and zeros elsewhere. The basic properties of this function can be easily proved using the identities in (\ref{eqn:pequiv_props}).
\begin{equation}
  \begin{aligned}
    C^Tofcol(x_{col}) = x_{col} \\
    R^Tofcol(x_{col}) = Ax_{col} + b \\
    R^Tofcol(x_{col}) = AC^Tofcol(x_{col}) + b
  \end{aligned}
\end{equation}

@@

% The type \verb|fin (m + n)| is the type of nonnegative integers less than \verb|m + n|. The \verb|row_indices| and \verb|column_indices| fields of the structure are vectors of length \verb|m| and \verb|n| respectively, containing elements of \verb|fin (m + n)|. The final field of the structure stipulates that all element of \verb|fin (m + n)| are an element of either \verb|row_indices| or \verb|col_indices|. This also implies, via a pigeonhole argument, that the two vectors are disjoint and have no duplicates. Given vectors $r$ and $c$ of row and column indices, it is possible to define the following minors of any vector $x \in \mathbb{Q}^{m + n}$.
% \begin{equation}\label{eqn:minor}
% \begin{aligned}
% x_{row} := (x_{row_0}, ... , x_{row_{m-1}}) \\
% x_{col} := (x_{col_0}, ... , x_{col_{n-1}})
% \end{aligned}
% \end{equation}

@@

% These ``row'' variables are then expressed as affine combinations of the ``column'' variables by the tableau. Given an $m \times n$ matrix $A$, and a constant vector $k \in \mathbb{Q}^m$, we can express the following relation between the current row variables $x_{row}$ and column variables $x_{col}$.
% \begin{equation}\label{eqn:axk}
%   x_{row} = Ax_{col} + k
% \end{equation}

@@

% \verb|T.to_partition.rowp.to_matrix| is an $m \times (m + n)$ matrix, which corresponds to the linear map taking the minor $x_{row}$ of $x$ defined in (\ref{eqn:minor}). \verb|T.to_partition.colp.to_matrix| does the same for the column variables. This equation is the same as the equation in (\ref{eqn:axk}). The notation \lstinline|⬝| is used for matrix multiplication.
